---
title: "Writing a Custom MCMC"
author: "NIMBLE Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{css, echo = FALSE}
h1, h4.author, h4.date {
  text-align: center;
}
```

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr_opts <- list(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  dpi = 300,
  out.width = "700px",
  fig.asp = 1 / 1.618,
  cache = FALSE,
  autodep = TRUE,
  cache.comments = TRUE,
  fig.align = "center",
  echo = TRUE,
  results = 'hide',
  eval = FALSE
)
do.call(knitr::opts_chunk$set, knitr_opts)
```

## Understanding the Random Walk Sampler (Metropolis Hastings)

The basic idea of a random walk sampler, is that we can draw samples from the posterior of $\theta$ by randomly choosing new values based on randomly changing the value of $\theta$ conditioned on the other parameters in the value. We accept or reject our new $\theta$ based on an accept ratio.

Here we will describe the basic algorithm assuming that the chain is initialized at $\theta_0$. For some data $y$, 

Algorithm: For MCMC iteration i,
1. Compute the log density $\pi(y,\theta_i)$.

2. Propose a new value of $\theta$, $\theta^*$, from a proposal distribution $q(\theta^*|\theta_i)$. We typically assume that $q(\theta^*|\theta_i) = \mbox{Normal}(\theta_i, \sigma)$. In this case, we have to choose the scale parameter $\sigma$.

3. Calculate the acceptance ratio, 
\[
  \alpha = \frac{\pi(y,\theta^*)q(\theta_i|\theta^*)}{\pi(y,\theta_i)q(\theta_i|\theta^*)}.
\]
Note that when using a symmetric proposal distribution (e.g., a normal distribution), the proposal distributions cancel out.

4. Randomly draw a value $u \sim U(0,1)$, and accept the new value as $\theta_{i+1} = 
theta^*$, if $u < \alpha$, otherwise, $\theta_{i+1} = \theta$. Note that $\alpha$ can be bigger than 1. In that case, we accept 100% of the time.

5. Repeat steps 1-4 with $i = i+1$ until $i$ reaches the requested number of iterations.

## Example

Consider a very basic linear regression of cars data in R, where distance depends on speed.

```{r}
library(nimble)
## Understanding MCMC with Nimble:
code <- nimbleCode({
  ## Choose priors:
  b0 ~ dflat() # dunif(-Inf, Inf) flat prior for intercept.
  b1 ~ dnorm(0, tau = 0.1)  # vague prior for slope.
  sigma ~ dunif(0, 100) 
  
  for( i in 1:n ) {
    mu[i] <- b0 + b1*speed[i]
    dist[i] ~ dnorm(mu[i], sd = sigma)
  }
})

inits <- function(){ list(b0 = rnorm(1), b1 = rnorm(1), sigma = rgamma(1,1))}
constants <- list(n = nrow(cars), speed = cars$speed)

model <- nimbleModel(code = code, 
                     data = list(dist = cars$dist), 
                     constants = list(n = nrow(cars), speed = cars$speed), 
                      inits = inits())

cmodel  <- compileNimble(model)
conf <- configureMCMC(model)
## Build the MCMC
mcmc    <- buildMCMC(conf, monitors = c("b0",  "b1", "sigma")) ## Track key parameters
## Compile the MCMC samplers into C++
cmcmc   <- compileNimble(mcmc, project = model)

## Now run some iterations of the MCMC
cmcmc$run(1000)
samples <- as.matrix(cmcmc$mvSamples)
out <- coda::as.mcmc(samples)

plot(out[,'b0'])
```

Now let's write our own Metropolis Hastings sampler to provide to our MCMC. We need to setup a nimble function that has the model, and a target variable to sample. We may want to include the scale as input via control to allow the user to set it. We need to make sure we add `contains = sampler_BASE` in the function. If we accept the change, we copy over the changed model `model` to the current model `mvSaved`, otherwise we revert to the original model `mvSaved` copied back to `model`.

```{r, eval = FALSE}
My_RW_sampler <- nimbleFunction(
    contains = sampler_BASE,
    setup = function(model, mvSaved, target, control = list()) {
        scale <- extractControlElement(control, "scale", 1) ## Set default = 1.
        ## Which nodes do we need to update.
        calcNodes <- model$getDependencies(target)  
    },
    run = function() {
        ## Get Current Value and log density.
        currentValue <- model[[target]]
        currentLogProb <- model$getLogProb(calcNodes)
        
        ## Draw a new value
        proposedValue <- currentValue + rnorm(1, 0, sd = scale)

        ## Add it to the model (this caches it directly in model with <<-)
        model[[target]] <<- proposedValue

        ## Calculate new log density
        proposalLogProb <- model$calculate(calcNodes)
        
        log_accept_prob <- proposalLogProb - currentLogProb
        
        ## Can use decide(log_accept_prob) instead of drawing u directly.
        u <- runif(1,0,1)
        if(u < exp(log_accept_prob))
          accept <- TRUE
        else
          accept <- FALSE
        
        if(accept) {
            ## return new model (saved when running "calculate")
            copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
        } else {
            ## return original model (saved before updating)
            copy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
        }    
    },
    ## Ignore this for now.
    methods = list(reset = function() {
    })    
)

## Let's add it to the MCMC
conf$removeSamplers('b0')
conf$addSampler(target = 'b0', type = 'My_RW_sampler', control = list(scale = 1.2))
conf$printSamplers('b0')
mcmc <- buildMCMC(conf, monitors = c("b0",  "b1", "sigma")) ## Track key parameters
cmcmc   <- compileNimble(mcmc, project = model, resetFunctions = TRUE)

## Now run some iterations of the MCMC
cmcmc$run(1000)
samples <- as.matrix(cmcmc$mvSamples)
out <- coda::as.mcmc(samples)
plot(out[,'b0'])
```
